ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}

### 2021-06-29 manuscript/main.tex
update manuscript to 2nd revision

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
ProSolo models single cells jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered polymorphic sites.
As a consequence, ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.

@@

Thus, variant callers for whole-genome-amplified single-cell data need to account for both amplification errors and allelic bias, in addition to accounting for the site-specific variation.
However, state of the art single cell SNV callers routinely assume fixed global rates when modeling uneven allelic coverage (up to dropout) and amplification errors. 
For example, to reflect the amplification error rate, both MonoVar\cite{zafar_monovar:_2016} and SCcaller\cite{dong_accurate_2017} work with global false positive error rates for calling the presence of an alternative allele at a particular site.
This assumes that $\Phi29$ polymerase is agnostic to local template sequence context, although it is not\cite{de_paz_high-resolution_2018}.
Similarly, for modelling allele dropout, MonoVar\cite{zafar_monovar:_2016} and SCIPhI\cite{singer_single-cell_2018} assume that one rate applies globally (and across all cells).
This neglects that allele dropout, as the extreme case of uneven allele coverage, varies greatly along the genome and in particular also between cells, because their DNA is amplified separately.
Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
Two tools that more variably model allelic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
Both estimate the minor allele frequency from nearby germline heterozygous sites. 
However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
Thus, to the best of our knowledge, there is no statistical model that allows for local variation of bias and errors due to amplification, and for statistically sound false discovery control when calling and genotyping SNVs in single cells.

@@

\paragraph{A mechanistically motivated model of amplification bias, trained on data, gives realistic coverage-specific single-cell (genotype) likelihoods.}
To account for MDA amplification bias up to the complete dropout of individual alleles, we distinguish between two alternative allele frequencies:
\begin{enumerate*}[label={(\roman*)}]
  \item The theoretical underlying allele frequency at a site in a single cell: $\theta_s$.
   This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
   However, the ratio of reads harboring the different alleles from a single cell sequencing experiment does not reflect the true allele frequency, because of the biases induced by amplification. 
  Instead, the ratio of reads reflects
  \item the allele frequency after its distortion through amplification bias.
   For a site with total coverage of $l$ reads, of which $k$ reads bear the alternative allele, the formal definition of this measurable frequency is $\rho_s=\frac{k}{l}$.
\end{enumerate*}
To accurately quantify the uncertainty introduced by amplification bias, we consider

@@

\begin{table}[tbp]
 \caption{Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}. Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.}
 \label{tab:Lodato}
 \renewcommand{\arraystretch}{1.15}

 \begin{center}
  \begin{tabular}{lrr}
   parameter  & slope        & intersect   \\
   \toprule
   $\alpha$   & -0.000027183 & 0.068567471 \\
   $\beta$    & 0.007454388  & 2.367486659 \\
   \midrule
   $w$        & 0.000548761  & 0.540396786 \\
   $\alpha_1$ & 0.057378844  & 0.669733191 \\
   $\alpha_2$ & 0.003233912  & 0.399261625 \\
   \bottomrule
  \end{tabular}
 \end{center}
\end{table}

\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
This establishes a formal, statistical argument to why one should consider bulk experiments in single cell sequencing whenever possible and why the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.

@@

Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

@@

For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
For example, an error-free homozygous alternative site is defined by allele frequency likelihood estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
And the dropout of an alternative allele is defined across allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.

@@

\paragraph{Biologically relevant imputation based on the bulk sample.}
As argued above, a large enough sampling of the bulk cell population that the single cell comes from should contain the single cell's genotype at a particular site, unless this cell is genuinely the first cell to harbour a mutation at that site.
This bulk background sample can thereby render credibility to single cell variants with low coverage, while at the same time eliminating amplification errors in the single cell sample, as these will not exist in the bulk sample.
Interestingly, the bulk sample also provides a mechanism of biologically meaningful imputation at sites that have no read coverage in the single cell.
If imputation is desired for sites with no read coverage in the single cell, we set $\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) = \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac{1}{2}) = \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) = 1$, rendering all (unknown underlying) single cell genotypes equally likely.
Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will usually favor germline genotypes over any existing (lower frequency) somatic genotypes at a site.

@@

We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}. We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 5.4.0}) to implement the benchmarking workflows.
For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).

@@

\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\

@@

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  All panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B. \newline \footnotesize
  Threshold parameters:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}

@@

At the same time, this germline ground truth caveat also indicates that the recall of SCIPhI's sensitive mode and ProSolo's imputation mode will be an overestimate.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
Both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation.
However, at single cell sites where a somatic mutation has created a true alternative allele, but no coverage is provided, we expect that both methods are most likely to call the homozygous reference germline genotype.
In then comparing this to the germline genotype as the ground truth, these calls will be classified as true negatives even though they really constitute false negatives, thus artificially increasing recall.
While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:F_1-scores_Laehnemann2017} for reference).

For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at reduced precision.
This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, both will further increase SCIPhI's wall time.
In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.

@@

Interestingly, a bulk coverage as low as 4X already provides drastic performance improvements in ProSolo's model and allows for a meaningful control over the false discovery rate (Supplementary Figure~\ref{fig:downsampling}, but higher bulk coverage will be necessary to detect low-frequency somatic variants (Supplementary Section~\ref{sec:somatic-recall}).

@@

Only in the dataset with more cells and a bulk sample with a lower coverage, SCIPhI can reach a higher recall than ProSolo by integrating information across cells (but at the cost false positive somatic mutation calls).

@@

Thus, any addition of genomic coverage and cells will further increase SCIPhI's wall clock runtime, where adding cells could prove especially troubling: adding cells will exponentially grow the space of possible tree topologies to explore.

>>>>

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
ProSolo models a single cell jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered sites, when downstream tools cannot handle missing data.
ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to quantify, such as allele dropout.

@@

Thus, variant callers for whole-genome-amplified single-cell data need to account for both amplification errors and allelic bias, in addition to accounting for the site-specific variation.
However, state of the art single cell SNV callers routinely assume fixed global rates when modeling uneven allelic coverage (up to dropout) and amplification errors. 
For example, to reflect the amplification error rate, both MonoVar\cite{zafar_monovar:_2016} and SCcaller\cite{dong_accurate_2017} work with global false positive error rates for calling the presence of an alternative allele at a particular site.
This assumes that $\Phi29$ polymerase is agnostic to local template sequence context, although it is not\cite{de_paz_high-resolution_2018}.
Similarly, for modelling allele dropout, MonoVar\cite{zafar_monovar:_2016} and SCIPhI\cite{singer_single-cell_2018} assume that one rate applies globally (and across all cells).
This neglects that allele dropout, as the extreme case of uneven allele coverage, varies greatly along the genome and in particular also between cells, because their DNA is amplified separately.
Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
Two tools that more variably model allelic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
Both estimate the minor allele frequency from nearby germline heterozygous sites. 
However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
Thus, to the best of our knowledge, there is no statistical model that allows for both local variation of bias and errors due to amplification, and for statistically sound false discovery rate control when calling and genotyping SNVs in single cells.

@@

\paragraph{A mechanistically motivated model of amplification bias, trained on data, gives realistic coverage-specific single-cell (genotype) likelihoods.}
To account for MDA amplification bias up to the complete dropout of individual alleles, we distinguish between two alternative allele frequencies:
\begin{enumerate*}[label={(\roman*)}]
  \item The true (but usually unknown) underlying allele frequency at a site in a single cell: $\theta_s$.
   This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
   However, the ratio of reads harboring the different alleles from a single cell sequencing experiment does not reflect the true allele frequency, because of the biases induced by amplification. 
  Instead, the ratio of reads reflects
  \item the allele frequency after its distortion through amplification bias.
   For a site with total coverage of $l$ reads, of which $k$ reads bear the alternative allele, the formal definition of this measurable frequency is $\rho_s=\frac{k}{l}$.
\end{enumerate*}

The goal is to estimate the likelihood density across the three possible underlying allele frequencies in the single cell (we denote $\tilde{\theta}_s$ as the density estimate across all $\theta_s\in\{0.0,0.5,1.0\}$).
To accurately quantify the uncertainty introduced by amplification bias, we consider

@@

\begin{table}[tbp]
 \caption{
 Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
 The full precision as used in the implementation is given in Supplementary Table~\ref{tab:Lodato-full}
 }
 \label{tab:Lodato}
 \renewcommand{\arraystretch}{1.15}

 \begin{center}
  \begin{tabular}{lrr}
   parameter  & slope        & intersect   \\
   \toprule
   $\alpha$   & -0.00003 & 0.06857 \\
   $\beta$    & 0.00745  & 2.36749 \\
   \midrule
   $w$        & 0.00055  & 0.54040 \\
   $\alpha_1$ & 0.05738  & 0.66973 \\
   $\alpha_2$ & 0.00323  & 0.39926 \\
   \bottomrule
  \end{tabular}
 \end{center}
\end{table}

\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
This establishes a formal, statistical argument why one should consider bulk experiments in single cell sequencing whenever possible: the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell, in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
At the same time, the absence of a single cell candidate mutation in a bulk sample (with sufficient sequencing depth) increases the probability of an amplification error.
As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.

@@

Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.
Further, as for the single cell sample, when referring to the likelihood density estimates across all possible allele frequencies (as opposed to a particular allele frequency), we denote this with $\tilde{\theta}_b$.

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across all possible allele frequencies $\tilde{\theta}_s$ and $\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

@@

For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel; for discussion on the validity of this assumption, see Supplementary Section \ref{sec:event-space-def}).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
For example, an error-free homozygous alternative site is defined by allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
And the dropout of an alternative allele is defined across allele frequency likelihood density estimates $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.

@@

\paragraph{Biologically relevant imputation based on the bulk sample.}
As argued above, a large enough sampling of the bulk cell population that the single cell comes from should contain the single cell's genotype at a particular site, unless this cell is genuinely the first cell to harbour a mutation at that site.
This bulk background sample can thereby render credibility to single cell variants with low coverage, while at the same time eliminating amplification errors in the single cell sample, as these will not exist in the bulk sample.
Interestingly, the bulk sample also provides a mechanism of biologically meaningful imputation at sites that have no read coverage in the single cell.
If imputation is desired for sites with no read coverage in the single cell, we set $\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) = \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac{1}{2}) = \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) = 1$, rendering all (unknown underlying) single cell genotypes equally likely.
Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will favor germline genotypes over any existing (lower frequency) somatic genotypes at a site, unless such a somatic genotype is present in a majority of cells.
Thus, such an imputation carries the potential to introduce erroneous calls (especially when looking at subclonal somatic mutations), and we recommend to instead use downstream tools that can accommodate for missing data and data uncertainty wherever they are available.

@@

We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}. We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 6.3.0}) to implement the benchmarking workflows.
For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).

@@

\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from SRA project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\

@@

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  All panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B. \newline \footnotesize
  Threshold parameters (not comparable across tools):
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}

@@

At the same time, this germline ground truth caveat also indicates that the recall of SCIPhI's sensitive mode and ProSolo's imputation mode will be an overestimate.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
Both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation.
However, at single cell sites where a somatic mutation has created a true alternative allele, but no coverage is provided, we expect that both methods are most likely to call the homozygous reference germline genotype.
In then comparing this to the germline genotype as the ground truth, these calls will be classified as true negatives even though they really constitute false negatives, thus artificially increasing recall.
While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:per-cell_prec-rec_Laehnemann2017} for reference).

For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at reduced precision.
This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, both will further increase SCIPhI's wall time.
In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores super-exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.

@@

Interestingly, a bulk coverage as low as 4X already provides drastic performance improvements in ProSolo's model and allows for a meaningful control over the false discovery rate (Supplementary Figure~\ref{fig:downsampling}; but higher bulk coverage will be necessary to detect low-frequency somatic variants, Supplementary Section~\ref{sec:somatic-recall}).

@@

Only in the dataset with more cells and a bulk sample with a lower coverage, SCIPhI can reach a higher recall than ProSolo by integrating information across cells (but at the cost of false positive somatic mutation calls).

@@

Thus, any addition of genomic coverage and cells will further increase SCIPhI's wall clock runtime, where adding cells could prove especially troubling: adding cells will super-exponentially grow the space of possible tree topologies to explore.
